---
title: "Case Study - BellaBeat and SmartWatch Analysis"
author: "Bryan Kim"
date: "2023-10-29"
output: html_document
resource_files:
- dailyActivity_merged.csv
- sleepDay_merged.csv
- Smart watch prices.csv
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing packages and opening libraries

```{r tidyverse}
install.packages("tidyverse")
install.packages("data.table")
library(tidyverse)
library(data.table)
```
## Importing Datasets
```{r}
activity <- read.csv("C:/Datasets/dailyActivity_merged.csv")
sleep <- read.csv("C:/Datasets/Sleepday_merged.csv")
prices <- read.csv("C:/Datasets/Smart watch prices.csv")
```
## Preview Datasets
```{r}
head(activity)
str(activity)

head(sleep)
str(sleep)

head(prices)
str(prices)
```
### Looking at dataset names to clean and rename
```{r}
colnames(activity)
colnames(sleep)
colnames(prices)
```
### Clean Names
```{r}
a <- rename_with(activity, tolower)
s <- rename_with(sleep, tolower)
p <- rename_with(prices, tolower)
```
#### Set Date Name and Time to same Format
```{r}
a <- a %>%
  rename(date = activitydate) %>%
  mutate(date = as_date(date, format = "%m/%d/%Y"))
s <- s %>%
  rename(date = sleepday) %>%
  mutate(date = as_date(date, format = "%m/%d/%Y %I:%M:%S %p"))
```
#### Ensure proper Date and Time Format
```{r}
head(a)
head(s)
```
## Verifying unique Id's & Smart Watch Models vs Price Due to Multiple Price Options
```{r}
n_distinct(a$id)
n_distinct(s$id)
n_distinct(p[2, 13])
```
## Clean Data, Looking for Duplicates
```{r}
sum(duplicated(a))
sum(duplicated(s))
sum(duplicated(p[2, 13]))
```
### Remove Duplicates, 0's, and Blank Rows in Datasets to remove non-usage and invalid data.
```{r}
a["totalsteps"][a["totalsteps"] == 0] <- NA
a_clean <- a %>%
  distinct() %>%
  drop_na()

s_clean <- s %>%
  distinct() %>%
  drop_na()

p_clean <- p[!apply(p == "", 1, all),]
p_clean <- p_clean %>%
  distinct() %>%
  drop_na()

p_clean <- unique(p_clean[ , c('brand', 'model', 'price..usd.')])
```
### Verify Duplicates got removed
```{r}
sum(duplicated(a_clean))
sum(duplicated(s_clean))
sum(duplicated(p_clean))
```
### Find out range of values and s.d. to reduce outliers
```{r}
range(a_clean$totalsteps)
range(s_clean$totaltimeinbed)
range(p_clean$price..usd.)
```
## Smartwatch Price for each Brand - Cleaning extra spaces and eliminating special characters
```{r}
p_trimmed <- p_clean %>%
  mutate(price..usd.=str_squish(price..usd.))

p_trimmed$price..usd. = as.numeric(gsub("[$,]", "", p_trimmed$price..usd.))

range(p_trimmed$price..usd.)
```

#### Find Interquartile range and Quartiles
```{r}
a_25 <- quantile(a_clean$totalsteps, .25)
a_75 <- quantile(a_clean$totalsteps, .75)
s_25 <- quantile(s_clean$totaltimeinbed, .25)
s_75 <- quantile(s_clean$totaltimeinbed, .75)
iqr_a <- IQR(a_clean$totalsteps)
iqr_s <- IQR(s_clean$totaltimeinbed)
```
#### Calculated lower and upper limits for outliers. However, decided to keep data due to not being able to determine if people slept an hour or 16 hours in a given day and there were values above the upper limit for activity from more active days.
```{r}
a_lower <- a_25 - 1.5*(iqr_a)
a_upper <- a_75 + 1.5*(iqr_a)
s_lower <- s_25 - 1.5*(iqr_s)
s_upper <- s_75 + 1.5*(iqr_s)

head(a_lower)
head(a_upper)
head(s_lower)
head(s_upper)
```

## Categorize amount of activity and sleep by each ID & rename the ID's from 1
```{r}
a_s <- merge(a_clean, s_clean, by = c("id", "date"))
a_s <- a_s %>% group_by(id) %>% mutate(id = cur_group_id())

head(a_s)
```

## Finding averages for Steps/Day vs Calories/Day of Smart Watch Prices by Brand
```{r}
a_s_avg <- a_s %>%
  group_by(id) %>%
  summarize(steps_avg = mean(totalsteps), sleep_avg = mean(totalminutesasleep), sedentary_avg = mean(sedentaryminutes), calories_avg = mean(calories), daily_use = sum(n()))

head(a_s_avg)

p_avg <- p_trimmed %>%
  group_by(brand) %>%
  summarize(avg_price = mean(price..usd.))

head(p_avg)
```

## Categorizing SmartWatches by less than, more than, or equal to Bellabeat Ivy ($249)
```{r}
p_avg_brand <- p_avg %>%
  select(brand, avg_price) %>%
  mutate(p_avg_brand = case_when(
    avg_price < 249 ~ "Less than Ivy",
    avg_price == 249 ~ "Same as Ivy",
    avg_price > 249 ~ "More than Ivy"
  ))

head(p_avg_brand)
```
## Grouping brands by prices
```{r}
p_avg_brand_percent <- p_avg_brand %>%
  group_by(p_avg_brand) %>%
  summarise(total = n()) %>%
  mutate(sum = sum(total)) %>%
  group_by(p_avg_brand) %>%
  summarise(percent = total/sum) %>%
  mutate(labels = scales::percent(percent))

p_avg_brand_percent$p_avg_brand <- factor(p_avg_brand_percent$p_avg_brand, levels = c("More than Ivy", "Same as Ivy", "Less than Ivy"))

head(p_avg_brand_percent)
```

## Total Minutes 
```{r}
a_s_avg %>%
    group_by(id, steps_avg) %>%
    ggplot(aes(id, steps_avg)) +
    geom_col(fill = "#f38488") + 
    geom_hline(yintercept = 7500) +
    labs(title = "Daily Steps", x = "User Id", y = "Average Steps") +
    theme(plot.title = element_text(hjust = 0.5))
a_s_avg %>%
    group_by(id, sleep_avg) %>%
    ggplot(aes(id, sleep_avg)) +
    geom_col(fill = "#344d70") + 
    geom_hline(yintercept = 480) +
    labs(title = "Daily Sleep", x = "User Id", y = "Average Asleep") +
    theme(plot.title = element_text(hjust = 0.5))
a_s_avg %>%
    group_by(id, sedentary_avg) %>%
    ggplot(aes(id, sedentary_avg)) +
    geom_col(fill = "#85c0cb") + 
    geom_hline(yintercept = 720) +
    labs(title = "Daily Minutes Sedentary", x = "User Id", y = "Average Minutes") +
    theme(plot.title = element_text(hjust = 0.5))
a_s_avg %>%
    group_by(id, calories_avg) %>%
    ggplot(aes(id, calories_avg)) +
    geom_col(fill = "#afa8ed") + 
    geom_hline(yintercept = 2000) +
    labs(title = "Daily Calories Burned", x = "User Id", y = "Average Calories") +
    theme(plot.title = element_text(hjust = 0.5))
a_s_avg %>%
    group_by(id, daily_use) %>%
    ggplot(aes(id, daily_use)) +
    geom_col(fill = "#cfd7e1") + 
    geom_hline(yintercept = 15) +
    labs(title = "Daily Device Usage", x = "User Id", y = "Number of Days") +
    theme(plot.title = element_text(hjust = 0.5))
```
## Correlation between Steps & Sedentary vs Calories.
```{r}
ggplot(a_s_avg, aes(x=steps_avg, y=calories_avg))+
  geom_smooth(color = "blue") + 
  labs(title = "Steps vs Average Calories Burned", x = "Average Steps", y= "Calories Burned") +
   theme(plot.title = element_text(hjust = 0.5, size=12))

ggplot(a_s_avg, aes(x=sedentary_avg, y=calories_avg))+
  geom_smooth(color = "blue") + 
  labs(title = "Sedentary vs Calories Burned", x = "Average Time Sedentary", y= "Calories") +
   theme(plot.title = element_text(hjust = 0.5, size=12))
```

```{r}
p_avg %>%
    group_by(brand, avg_price) %>%
    ggplot(aes(brand, avg_price)) +
    geom_col(fill = "#018704") + 
    geom_hline(yintercept = 249) +
    labs(title = "Cost of Smart Watches", x = "Brand", y = "Average Price (USD)") +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))
p_avg_brand_percent %>%
    group_by(p_avg_brand, percent) %>%
    ggplot(aes(x = "", y = percent, fill = p_avg_brand)) +
    geom_col(color = "black") +
    geom_text(aes(label = labels), position = position_stack(vjust = 0.5)) +
    labs(title = "Cost of Smart Watches by Percent") +
    labs(fill = "Cost vs Ivy\n") +
    coord_polar("y") +
    theme_bw() + 
    theme(axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, size=12, face = "bold"))
```